---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Get economic tax data

```{r}
#| label: includes
#| include: false

source(here::here("r", "libraries.r"))
source(here::here("r", "libraries_ts.r"))
source(here::here("r", "constants.r"))
source(here::here("r", "functions.r"))

```

```{=html}
<!-- links


https://dol.ny.gov/occupational-and-industry-data


-->
```

## The New York State Department of Labor's QCEW data

The New York State Department of Labor (NYSDOL) provides a version of the BLS Quarterly Census of Employment and Wages (QCEW) that has quarterly county-level wage, establishment, and employment data in a format we need. The data currently available from 2000 through the fourth quarter of 2022.

```{r}
#| label: ONETIME-get-save-nysdol-qcew
#| eval: false
#| include: false

files <- fs::dir_ls(path=here::here("data", "nyqcew"),
                    glob = "*csv")
files

df <- vroom(files)
glimpse(df)
names(df)
# [1] "AREATYPE"    "AREA"        "YEAR"        "QUARTER"     "NAICS"       "NAICS_LEVEL" "NAICS_TITLE" "ESTAB"       "MNTH1EMP"    "MNTH2EMP"   
# [11] "MNTH3EMP"    "TOTWAGE"     "Owner"  
count(df, AREATYPE) # County, State
count(df |> filter(AREATYPE=="County"), AREA)
count(df, Owner)
# Owner                              n
# <chr>                          <int>
# 1 Federal Government            102215
# 2 Local Government              245916
# 3 Private                      7052970
# 4 State Government              134571
# 5 Total Government              558464
# 6 Total Private and Government 8068156


df1 <- df |> 
  rename_with(tolower) |> 
  filter(areatype %in% c("County", "State"),
         owner %in% c("Total Private and Government", "Local Government", "Federal Government")) |> 
  filter(naics=="00" | (naics_level==4 & naics=="6111"))

saveRDS(df1, here::here("data", "nyqcew", "mtaqcew_raw.rds"))

```


```{r}
#| label: ONETIME-save-nysdol-qcew-slim
#| eval: false
#| include: false


mtawage1 <- readRDS(here::here("data", "nyqcew", "mtaqcew_raw.rds"))
count(mtawage1, year, quarter) |> ht()

mtawage2 <- mtawage1 |> 
  filter(area=="New York State" | area %in% constants$mtaregion$area) |> 
  mutate(date=yq(paste0(year, "-", quarter)),
         areatype=str_to_lower(areatype),
         region=case_when(area %in% mtaregion$area[1:5] ~ "nyc",
                          area %in% mtaregion$area[6:12] ~ "suburban",
                          area == "New York State" ~ "nys",
                          TRUE ~ "ERROR"),
         owncode=factor(owner, 
                        levels=c("Total Private and Government", "Local Government", "Federal Government"),
                        labels=c("total", "locgov", "fedgov")),
         ind=case_when(naics=="00" ~ "allind",
                       naics=="6111" ~ "k12"))
count(mtawage2, areatype, region, area)
count(mtawage2, owncode, owner)
count(mtawage2, ind, naics, naics_level, naics_title)

mtawage3 <- mtawage2 |> 
  select(date, year, qtr=quarter, region, area, ind, owncode, estab:totwage) |> 
  pivot_longer(cols=estab:totwage)

saveRDS(mtawage3, here::here("data", "nyqcew", "mtaqcew_slim.rds"))

```


```{r}
#| label: ONETIME-mta-wagebase
#| eval: false
#| include: false

mtawage3 <- readRDS(here::here("data", "nyqcew", "mtaqcew_slim.rds"))

mtawage3 |> 
  filter(area=="Dutchess County",
         ind=="allind",
         name=="totwage") |> 
  filter(year(date)>=2010) |> 
  ggplot(aes(date, value)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  scale_x_date(breaks = "3 months") +
  facet_wrap(~owncode, scales="free_y", ncol=1) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5)) 

check <- mtawage3  |> filter(area=="New York County", date==max(date))
check <- mtawage3  |> filter(region=="nyc", date==max(date))

xk12 <-  mtawage3 |> 
  filter(ind=="k12" & owncode=="locgov")

xk12all <-  mtawage3 |> 
  filter(ind=="k12")
xk12all |> filter(region=="nys", date==max(date))
xk12all |> filter(region=="nyc", date==max(date)) # no local gov k12 data for nyc -- so must subtract the total owner

xfed <- mtawage3 |> 
  filter(ind=="allind" & owncode=="fedgov")
xfed |> filter(region=="nys", date==max(date))
xfed |> filter(region=="nyc", date==max(date))

count(xk12, area)


mtabase1 <- mtawage3 |> 
  filter(ind=="allind", owncode=="total") |> 
  select(-c(ind, owncode)) |> 
  rename(total=value) |> 
  left_join(xk12all |> 
              filter(owncode=="total") |> # use this due to data limitations
              select(date, area, name, xk12=value),
            by = join_by(date, area, name)) |> 
  left_join(xfed |> 
              select(date, area, name, xfed=value),
            by = join_by(date, area, name)) |> 
  mutate(pmtbase=total - naz(xk12) - naz(xfed))

count(mtabase1, area, region)
skim(mtabase1)

saveRDS(mtabase1, here::here("data", "mta_wagebase.rds"))


```

Here is a summary of the resulting data file.


```{r}
#| label: summary
#| eval: true
#| include: true

mtabase1 <- readRDS(here::here("data", "mta_wagebase.rds"))
skim(mtabase1)

```

And correlation between total wages and the tax base after removing federal wages and K12 wages:

```{r}
#| label: correlation
#| eval: true
#| include: true

mtabase1 |> filter(name=="totwage") |> select(total, pmtbase) |> cor()

```



```{r}
#| label: explore
#| eval: false
#| include: false

mtabase1 <- readRDS(here::here("data", "mta_wagebase.rds"))

mtabase1 |> filter(name=="totwage") |> select(total, pmtbase) |> cor()

mtabase1 |> 
  filter(name=="totwage") |> 
  summarise(total=sum(total), .by=c(date, region)) |> 
  ggplot(aes(date, total)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  scale_x_date(breaks = "3 months") +
  facet_wrap(~region, scales="free_y", ncol=1) +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5)) 


mtabase1 |> 
  filter(name=="totwage") |> 
  summarise(total=sum(total), .by=c(date, region)) |> 
  arrange(date) |> 
  mutate(pch=total / lag(total) - 1, .by=region) |> 
  mutate(region=factor(region, levels=c("nyc", "suburban", "nys"))) |> 
  filter(year(date) >= 2010) |> 
  ggplot(aes(date, pch)) +
  geom_line() +
  geom_point() +
  geom_smooth() +
  geom_hline(yintercept = 0) +
  scale_x_date(name=NULL, breaks = "3 months") +
  scale_y_continuous(name="% change from year ago", 
                     breaks=seq(-.5, .5, .05),
                     labels=label_percent(accuracy=1)) +
  facet_wrap(~region, ncol=1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = -90, vjust = 0, hjust=0.5))


```


